A Linear Solution of the Linearity Challenge

### 2025-12-13 paper/Examples.lagda.tex
restored examples

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Sum hiding (reduce; swap)
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.Nat using (ℕ; zero; suc)
open import Data.Fin using (zero; suc; #_)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)
open import Function using (_$_)
open import Data.Maybe

@@

In this section we illustrate a few simple examples of processes. We start by
encoding the type of a boolean value being produced as a choice between ``going
left'' (which we interpret as true) and ``going right'' (which we interpret as
false).

@@

\begin{code}
true : Process [ 𝔹 ]
true = left (< ≫) close

false : Process [ 𝔹 ]
false = right (< ≫) close
\end{code}

@@

\begin{code}
if_else : ∀{Γ} → Process Γ → Process Γ → Process (dual 𝔹 ∷ Γ)
if P else Q = case (< ≫) (wait (< ≫) P) (wait (< ≫) Q)
\end{code}

A term $\AgdaFunction{if}~P~\AgdaFunction{else}~Q$ is a process that waits for a
boolean value (\cf the $\AgdaFunction{dual}~\mathbb{B}$ type at the front of its
typing context) and continues as either $P$ or $Q$ depending on whether it
receives true or false.
%
We can then define a process $\AgdaFunction{drop}~P$ that waits for consumes a
boolean and continues as $P$ regardless of its value.

\begin{code}
drop : ∀{Γ} → Process Γ → Process (dual 𝔹 ∷ Γ)
drop P = if P else P
\end{code}

@@

\begin{code}
!!_ : Process [ 𝔹 ] → Process [ 𝔹 ]
!!_ B = cut ≫ B (if false else true)

_&&_ _||_  : Process [ 𝔹 ] → Process [ 𝔹 ] → Process [ 𝔹 ]
A && B   = cut ≫ A (cut ≫ B (if (link (< ≫)) else (drop false)))
A || B   = !! ((!! A) && (!! B))
\end{code}

Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
produces a boolean value with a process that consumes a boolean value and the
use of \AgdaInductiveConstructor{link} for copying a boolean value from one
channel to another. 
%
Here we have chosen to define the disjunction \AgdaFunction{||} from the
conjunction \AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's
laws.

If we wish to test our definitions, we can implement a simple (lazy) evaluator
using the deadlock freedom property. We have not proved a termination result,
but since linear logic enjoys cut elimination we can safely annotate the
evaluator as terminating.

\begin{code}
{-# TERMINATING #-}
eval : ∀{Γ} → Process Γ → Process Γ
eval P with deadlock-freedom P
... | inj₁ (Q , _ , _)  = Q
... | inj₂ (Q , _)      = eval Q
\end{code}

Now if we ask Agda to normalize the goal
$\AgdaFunction{eval}~(\AgdaFunction{false}~\AgdaFunction{||}~\AgdaFunction{false})$
we obtain
$\AgdaInductiveConstructor{right}~(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{•})~\AgdaInductiveConstructor{close}$,
that is the definition of \AgdaFunction{false}, as expected.

@@

Below we define a persistent, polymorphic echo \AgdaInductiveConstructor{server}
that waits for a value and sends it back unchanged. The server is polymorphic in
the sense that the type of the value is arbitrary and denoted by the type
variable $X$. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
library and creates an element of \AgdaDatatype{Fin} from the natural number
$n$. The function \AgdaFunction{\$} is also defined in the standard library and
denotes function application. We use it here as ``command separator'' to save a
few parentheses.

\begin{code}
echo :  let X = var (# 0) in
        Process [ `! (`∀ (X ⊸ X)) ]
echo =  server (< ≫) un-[] $
        all (< ≫) λ X →
        join (< ≫) $
        link (< ≫)
\end{code}

@@

\begin{code}
echo-true : Process [ 𝔹 ]
echo-true = cut ≫ echo (client (< ≫) $
                       ex (< ≫) $
                       fork (< ≫) ≫ true (link (< ≫)))
\end{code}

If we ask Agda to evaluate this process we obtain
$\AgdaInductiveConstructor{left}~(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{•})~\AgdaInductiveConstructor{close}$,
that is the definition of \AgdaFunction{true} as expected.

@@

% \begin{code}
% ⊗-comm : let X = var (# 1) in
%          let Y = var (# 0) in
%          Process [ `∀ (`∀ ((X ⊗ Y) ⊸ (Y ⊗ X))) ]
% ⊗-comm = all (< ≫) λ X →
%          all (< ≫) λ Y →
%          join (< ≫) $
%          join (> < ≫) $
%          fork (> > < ≫) (< ≫)
%               (link (< ≫))
%               (link (< ≫))
% \end{code}

>>>>

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Function using (_$_)
open import Data.Sum using (inj₁; inj₂)
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.Fin using (zero; suc; #_)
open import Data.List.Base using ([]; _∷_; [_])
open import Relation.Unary

@@

In this section we illustrate a few simple examples of processes encoded in our
formalization. We start by encoding the type of a boolean value being produced
as a choice between ``going left'' (which we interpret as true) and ``going
right'' (which we interpret as false).

@@

\begin{code}
true : Proc [ 𝔹 ]
true = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))

false : Proc [ 𝔹 ]
false = select (ch ⟨ < ≫ ⟩ inj₂ (close ch))
\end{code}

@@

\begin{code}
if_else : ∀{Γ} → Proc Γ → Proc Γ → Proc (dual 𝔹 ∷ Γ)
if P else Q = curry∗ case ch (< ≫) ( wait (ch ⟨ < ≫ ⟩ P)
                                   , wait (ch ⟨ < ≫ ⟩ Q))
\end{code}

A term $\AgdaFunction{if}~P~\AgdaFunction{else}~Q$ is a process that waits for a
boolean value (\cf the $\AgdaFunction{dual}~\mathbb{B}$ type at the front of its
typing context) and continues as either $P$ or $Q$ depending on whether it
receives true or false. We use \AgdaFunction{curry∗} (defined in
\Cref{sec:context-agda}) to curry the constructor
\AgdaInductiveConstructor{case} so that we can supply its arguments one by one
saving a few parentheses and reducing clutter.

Next we define a process $\AgdaFunction{drop}~P$ that consumes a boolean and
continues as $P$ regardless of its value.

\begin{code}
drop : ∀{Γ} → Proc Γ → Proc (dual 𝔹 ∷ Γ)
drop P = if P else P
\end{code}

@@

\begin{code}
!!_ : Proc [ 𝔹 ] → Proc [ 𝔹 ]
!!_ B = curry∗ cut B ≫ (if false else true)

_&&_ _||_  : Proc [ 𝔹 ] → Proc [ 𝔹 ] → Proc [ 𝔹 ]
A && B   =  curry∗ cut A ≫ $
            curry∗ cut B ≫ $
            if (curry∗ link ch (< ≫) ch) else (drop false)
A || B   = !! ((!! A) && (!! B))
\end{code}

The function \AgdaFunction{\$} is defined in Agda's standard library and denotes
function application. We use it here as ``separator'' to save a few parentheses.
%
Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
produces a boolean value with a process that consumes a boolean value and the
use of \AgdaInductiveConstructor{link} for copying a boolean value from one
channel to another. 
%
Here we have chosen to define the disjunction \AgdaFunction{||} from the
conjunction \AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's
laws.

If we wish to test our definitions, we can implement a simple evaluator using
the deadlock freedom property. We have not proved a termination result, but
since linear logic enjoys cut elimination we can safely annotate the evaluator
as terminating.

\begin{code}
{-# TERMINATING #-}
eval : ∀{Γ} → Proc Γ → Proc Γ
eval P with deadlock-freedom P
... | inj₁ (Q , _ , _)  = Q
... | inj₂ (Q , _)      = eval Q
\end{code}

Now if we ask Agda to normalize the goal
$\AgdaFunction{eval}~(\AgdaFunction{false}~\AgdaFunction{||}~\AgdaFunction{false})$
we obtain
$\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch~⟨~<~•~⟩~inj₂}~(\AgdaInductiveConstructor{close~ch}))$,
that is the definition of \AgdaFunction{false}, as expected.

@@

Below we define a persistent, polymorphic echo \AgdaInductiveConstructor{server}
that waits for a value and sends it back unchanged. The server is polymorphic in
the sense that the type of the value is arbitrary and denoted by the type
variable $X$. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
library and creates an element of \AgdaDatatype{Fin} from the natural number
$n$.

\begin{code}
echo : let X = var (# 0) in
       Proc [ `! (`∀ (X ⊸ X)) ]
echo = curry∗ server ch (< ≫)
             ( un-[]
             , curry∗ all ch (< ≫) λ X →
               curry∗ join ch (< ≫) $
               curry∗ link ch (< ≫) ch)
\end{code}

@@

\begin{code}
echo-true : Proc [ 𝔹 ]
echo-true = curry∗ cut echo ≫ $
            curry∗ client ch (< ≫) $
            curry∗ ex ch (< ≫) $
            curry∗ fork ch (< ≫) $ true ⟨ ≫ ⟩ curry∗ link ch (< ≫) ch
\end{code}

If we ask Agda to evaluate this process we obtain \AgdaFunction{true}, as
expected.

@@
